---
title: "Getting Started with ausOpenData"
output: rmarkdown::html_vignette
author: "Christian Martinez"
vignette: >
  %\VignetteIndexEntry{Getting Started with ausOpenData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(ausOpenData)
library(ggplot2)
library(dplyr)
```

## Introduction

Welcome to the `ausOpenData` package, an R package dedicated to helping R users connect to the [Austin Open Data Portal](https://data.austintexas.gov/)!

The `ausOpenData` package provides a streamlined interface for accessing Austin' vast open data resources. It connects directly to official City of Austin open data portals, including datasets hosted across Socrata-powered city domains, helping users bridge the gap between raw city APIs and tidy data analysis. This package is part of a broader ecosystem of open data tools designed to provide a consistent interface across cities. It does this in two ways:

### The `aus_pull_dataset()` function

The primary way to pull data in this package is the `aus_pull_dataset()` function, which works in tandem with `aus_list_datasets()`. You do not need to know anything about API keys or authentication.

The first step would be to call the `aus_list_datasets()` to see what datasets are in the list and available to use in the `aus_pull_dataset()` function. This provides information for thousands of datasets found on the portal.

```{r aus-list-datasets}
aus_list_datasets() |> head()
```

The output includes columns such as the dataset title, description, and link to the source. The most important fields are the dataset `key` and `id`. You need **either** in order to use the `aus_pull_dataset()` function. You can put **either** the key value or id value into the `dataset =` filter inside of `aus_pull_dataset()`.

For instance, if we want to pull the dataset `Golf Courses in Austin`, we can use either of the methods below:

```{r aus-vacant-pull}
aus_golf_courses <- aus_pull_dataset(
  dataset = "dtkn-v97q", limit = 2, timeout_sec = 90)

aus_golf_courses <- aus_pull_dataset(
  dataset = "golf_courses_in_austin", limit = 2, timeout_sec = 90)
```

No matter if we put the `uid` or the `key` as the value for `dataset =`, we successfully get the data!

### The `aus_any_dataset()` function

The easiest workflow is to use `aus_list_datasets()` together with `aus_pull_dataset()`.

In the event that you have a particular dataset you want to use in R that is not in the list, you can use the `aus_any_dataset()`. The only requirement is the dataset’s API endpoint (a URL provided by the Austin Open Data portal). Here are the steps to get it:

1. On the Austin Open Data Portal, go to the dataset you want to work with.
2. Click on "Export" (next to the actions button on the right hand side).
3. Click on "API Endpoint".
4. Click on "SODA2" for "Version".
5. Copy the API Endpoint.

Below is an example of how to use the `aus_any_dataset()` once the API endpoint has been discovered, that will pull the same data as the `aus_pull_dataset()` example:

```{text}
aus_golf_courses <- aus_any_dataset(json_link = "https://data.austintexas.gov/resource/dtkn-v97q.json", limit = 2)
```

### Rule of Thumb

While both functions provide access to Austin Open Data, they serve slightly different purposes.

In general:

- Use `aus_pull_dataset()` when the dataset is available in `aus_list_datasets()`
- Use `aus_any_dataset()` when working with datasets outside the catalog

Together, these functions allow users to either quickly access the datasets or flexibly query any dataset available on the Austin Open Data portal.

## Real World Example

Each food establishment in Austin has to be properly inspected, and the dataset with the inspection results is [found here](https://data.austintexas.gov/Health-and-Community-Services/Food-Establishment-Inspection-Scores/ecmv-9xxi/about_data). In R, the `ausOpenData` package can be used to pull this data directly.

By using the `aus_pull_dataset()` function, we can gather information on these food inspection results, and filter based upon any of the columns inside the dataset.

Let's take an example of 3 requests that are routine inspections. The `aus_pull_dataset()` function can filter based off any of the columns in the dataset. To filter, we add `filters = list()` and put whatever filters we would like inside. From our `colnames` call before, we know that there is a column called "city" which we can use to accomplish this.

```{r filter-routine}

aus_routine_results <- aus_pull_dataset(dataset = "ecmv-9xxi",limit = 3, timeout_sec = 90, filters = list(process_description = "Routine Inspection"))

aus_routine_results

# Checking to see the filtering worked
aus_routine_results |>
  distinct(process_description)
```

Success! From calling the `aus_routine_results` dataset we see there are only 3 rows of data, and from the `distinct()` call we see the only proccess featured in our dataset is "Routine Inspection".

One of the strongest qualities this function has is its ability to filter based off of multiple columns. Let's put everything together and get a dataset of *50* businesses that occur in Austin in council district 8.

```{r filter-aus-hundreds}
# Creating the dataset
aus_routine_100 <- aus_pull_dataset(dataset = "6rrh-rzua", limit = 50, timeout_sec = 90, filters = list(process_description = "Routine Inspection", score = "100"))

# Calling head of our new dataset
aus_routine_100 |>
  slice_head(n = 6)

# Quick check to make sure our filtering worked
aus_routine_100 |>
  summarize(rows = n())

aus_routine_100 |>
  distinct(process_description)

aus_routine_100 |>
  distinct(score)
```

We successfully created our dataset that contains 50 requests regarding the scores of 100 routine food inspections in Austin Texas.

## Summary

The `ausOpenData` package serves as a robust interface for the Austin Open Data portal, streamlining the path from raw city APIs to actionable insights. By abstracting the complexities of data acquisition—such as pagination, type-casting, and complex filtering—it allows users to focus on analysis rather than data engineering.

As demonstrated in this vignette, the package provides a seamless workflow for targeted data retrieval, automated filtering, and rapid visualization.

## How to Cite

If you use this package for research or educational purposes, please cite it as follows:

Martinez C (2026). ausOpenData: Convenient Access to Austin Open Data API Endpoints. R package version 0.1.0, <https://martinezc1.github.io/ausOpenData/>.
